---
title: "Great Lakes Fish"
author: Jennessa Brunette
format: html
editor: visual
execute: 
  keep-md: true
---

## Great Lake Fish Analysis

```{r}
#Load the tidyverse
library(tidyverse)
library(kableExtra)

```

```{r}
fishing <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-06-08/fishing.csv')
stocked <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-06-08/stocked.csv')

```

The `echo: false` option disables the printing of code (only output is displayed).

```{r}
#install.packages("tidymodels")
library(tidymodels)

my_data_splits <- initial_split(fishing, prop = 0.5)

exploratory_data <- training(my_data_splits)
test_data <- testing(my_data_splits)
```

# Introduction

The data I will be using is from Great Lakes Fishery Commission and was collected as early as 1867. The data had two sets of variables to look at: fishing and stocked data. This data includes 5 main variables from the fishing data set regarding the 51 Great Lakes fish species observed. The variables include year of measurement, species of fish, region of the U.S, lake name, and grand total of observed.

# Interesting Questions to ask

-   Which lake has the most variety of species?

-   How many of each fish species do we have by region?

-   What is the total amount of observed fish among the great lake fish species?

# Hypothesis

-   I hypothesize the average number of species will increase by year

-   I hypothesize the average number of species per lake will increase by year.

-   I hypothesize there will be a greater variety of species as time goes on.

This table shows the number of observed species. There are 51 species.

```{r}
exploratory_data %>%
  count(species)
```

```{r}
#install.packages("skimr")
library(skimr)
exploratory_data %>%
  skim()
```

```{r}
fishing %>%
  select (species, grand_total)
```

# Which lake has the most variety of species?

```{r}
fishing %>%
  group_by(lake, year, species) %>%
  summarize(total_observed = sum(grand_total, rm.na = TRUE)) %>%
  mutate(total_observed = ifelse(is.na(total_observed), 0, total_observed))%>%
  filter(total_observed >0)%>%
  summarize(number_species = n()) %>%
  ggplot() + geom_line(aes(x = year, y = number_species, color = lake))
```

This graph shows which lakes have more species than others. For example, lake Erie seemed to have the most number of species around 1995 but then during the 2000s experienced a decline. It now looks like during 2009 lake Erie and lake Huron had very similar number of species present.

In regards to my hypothesis this graph also provides insight into that. This graph tells us as time progresses, number of species in each lake also progress. However there are some exceptions where the number of species decline during a time period. For example, around 1895 lake superior experienced a decrease in number of species observed.

Overall from this graph I can accept my hypothesis that the number of species will increase over the years and there will be more variety in species. But, I cannot say that the species will increase every year, since the data here shows some years species experienced a decline.

In regards to the decline in species some years, I can assume this could be due to natural selection or over fishing.

# How many of each fish species do we have by region?

```{r}
fishing%>%
  group_by(species) %>%
  summarize(region)
fishing %>%
  group_by(lake) %>%
  summarize(region)
fishing %>%
  count(region)
```

```{r}
fishing%>%
ggplot() +
  geom_bar(aes(x = region),
               color = "pink", fill = "purple") +
  coord_flip()
  
```

This graph corresponds to the previous table. Here we can see the count of species by region. It looks like Canada has the highest count of 7479 fish. I assume this is because Canada's border runs along Lake superior, Huron, Ontario and Erie. Minnesota has the least amount of observed species.

# What is the total amount of observed fish among the great lakes fish species?

```{r}
fishing %>%
  ggplot() +
  geom_bar(mapping = aes(x = species)) +
  labs(title ="Counts of Fish Species",
       x = "Species", y = "Count") +
coord_flip()
```

This graph shows us visually how many of each fish species we have among the Great Lakes. From this graph I can tell Whitefish has the most species, followed by Walleye and then Northern Pike. Sheepshead has the lowest number of species.

I found this question interesting to ask because there is a lot of species of fish among the great lakes. It was helpful to see the species laid out and made me more aware of how many species of fish are in the Great Lakes.

# Future work

To test my hypothesis further I would create a sample population to see if certain fish species increased more successfully than others throughout the years. To see if there is a correlation between certain fish species and their ability to successfully increase I could use a distribution curve for each individual fish species.
