---
title: "Great Lakes Fish"
author: Jennessa Brunette
format: html
editor: visual
execute: 
  keep-md: true
---

## Great Lake Fish Analysis

```{r}
#Load the tidyverse
library(tidyverse)
library(kableExtra)

```

You can add options to executable code like this

```{r}
fishing <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-06-08/fishing.csv')
stocked <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-06-08/stocked.csv')

```

The `echo: false` option disables the printing of code (only output is displayed).

```{r}
#install.packages("tidymodels")
library(tidymodels)

my_data_splits <- initial_split(fishing, prop = 0.5)

exploratory_data <- training(my_data_splits)
test_data <- testing(my_data_splits)
```

# About Our Data

This data includes 5 main variables of the fishing data set regarding the 51 Great Lakes fish species observed. The variables include year of measurement, species of fish, region of the U.S, lake name, and grand total of observed.

# Interesting Questions to ask

-   Which lake has the most variety of species?

-   Will the species of fish change in each region every year?

-   How many of each fish species do we have? by region?

# Hypothesis

-   I hypothesize the average number of species will increase by year

-   I hypothesize the average number of species per lake will increase by year.

-   I hypothesize there will be a greater variety of species as time goes on.

This table shows the number of observed species. There are 51 species.

```{r}
exploratory_data %>%
  count(species)
```

```{r}
#install.packages("skimr")
library(skimr)
exploratory_data %>%
  skim()
```

```{r}
fishing %>%
  select (species, grand_total)
```

# Answering our questions & Hypothesis

```{r}
fishing %>%
  group_by(lake, year, species) %>%
  summarize(total_observed = sum(grand_total, rm.na = TRUE)) %>%
  mutate(total_observed = ifelse(is.na(total_observed), 0, total_observed))%>%
  filter(total_observed >0)%>%
  summarize(number_species = n()) %>%
  ggplot() + geom_line(aes(x = year, y = number_species, color = lake))
```

This graph shows which lakes have more species than others. For example, lake Erie seemed to have the most number of species around 1995 but then during the 2000s experienced a decline. It now looks like during 2009 lake Erie and lake Huron had very similar number of species present.

In regards to my hypothesis this graph also provides insight into that. This graph tells us as time progresses, number of species in each lake also progress. However there are some exceptions where the number of species decline during a time period. For example, around 1895 lake superior experienced a decrease in number of species observed.

Overall from this graph I can accept my hypothesis that the number of species will increase over the years and there will be more variety in species. But, I cannot say that the species will increase every year, since the data here shows some years species experienced a decline.

In regards to the decline in species some years, I can assume this could be due to natural selection or over fishing.

# Number of species per year in Lake Erie

```{r}
fishing %>%
  filter(lake == "Erie")
```

This table shows the variety of species observed in lake Erie by year.

# How many of each species do we have by region?

```{r}
fishing%>%
  group_by(species) %>%
  summarize(region)
fishing %>%
  group_by(lake) %>%
  summarize(region)
fishing %>%
  count(region)
```

```{r}
fishing%>%
ggplot() +
  geom_bar(aes(x = region),
               color = "pink", fill = "purple") +
  coord_flip()
  
```

This graph corresponds to the previous table. Here we can see the count of species by region. It looks like Canada has the highest count of 7479 fish. I assume this is because Canada's border runs along Lake superior, Huron, Ontario and Erie.

```{r}
fishing %>%
  ggplot() +
  geom_bar(mapping = aes(x = species)) +
  labs(title ="Counts of Fish Species",
       x = "Species", y = "Count") +
coord_flip()
```

This graph shows us visually how many of each fish species we have. From this graph I can tell Whitefish has the most species, followed by Walleye and then Northern Pike. Sheepshead has the lowest number of species.

I found this question interesting to ask because there is a lot of species of fish among the great lakes. It was helpful to see the species laid out and made me more aware of how many species of fish are in the Great Lakes
